Experiences Gamifying Developer Adoption\\ of Practices and Tools

### 2013-10-23 paper_id1.tex
first submitted version

\title{Experiences Gamifying Developer Practices}
%\subtitle{}
% 
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.   We tested an idea with industrial software developers that adding game-like feedback to the development environment would improve adoption of tools and practices for code navigation.  We conducted this study in three stages: a pre-study to validate our ideas, pilot of a tool with some game features to motivate better navigation practices, and feedback interviews for a post-pilot survey.  We found that a software engineering game with competitive elements influences some developers towards changing their practices.
%We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
%
% In addition to training, discussing, and presenting software engineering practices and tools, we demonstrate a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).  
%
%We determine through a survey of 130 developers the things we can do to motivate them to adopt 
%Blaze monitors developer practice and tool use over the long term and reports the lasting change a tool deployment or communication effort makes in everyday habits. 
%
% Embedding Blaze in the developer's work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.   Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.  
% With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.  
%
%Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.  We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools. The Blaze tool was well received by most participants, however, navigation practices were not significantly affected by the Blaze feedback mechanism.  We conclude that more proactive interventions are necessary to encourage developers to change their code navigation habits.
\end{abstract}

@@

Sascha is a developer with years of experience maintaining code who knows how to navigate through code fairly well.  Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE.  When Sascha's company started a game around code navigation in Visual Studio, he became interested in being a part of the fun.  The game gave bonus points for using Sando, a new code search tool, so he started using it instead of some of the built-in search commands.  Sascha soon topped the game's leaderboard and discovered he became motivated to try the next practices included in the game.

Like Sascha, software developers who maintain code face additional challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Over the years, developers learn techniques to analyze code and locate relevant areas for an enhancement.  As new practices and tools get introduced, developers may not learn about them unless they interact with a peer developer who uses them\cite{wbsnipes:Hill2011Peer}.    
%For instance, the commands they use to navigate the code base in the development environment may make them less efficient than using newer practices or tools.  Robillard et al. show that developers who use commands that allow them to follow code structure are more effective and efficient than developers who browse through code \cite{wbsnipes:Robillard2004How}.

Enter the toolsmith,  who responds to developers' challenges by creating tools such as code search tools that make the job easier and more efficient.  The toolsmith needs to spread the word about advanced tools they develop and determine their effect on developer practices over the long term.  Is there something we can do to encourage developers to adopt new practices and tools that also provides feedback to developers and toolsmiths on the resulting improvements?

One option for motivating people to adopt new practices is to apply elements of game mechanics or ``gamify'' the practice. Gamification is the incorporation of game elements into an activity that people do not typically consider a game\cite{2013Oxford}.  We envisioned a tool in our prior work to motivate developers by providing constant feedback and allowing them to compare their improvements to how their peers are doing  \cite{Snipes2013Towards}.  This tool called Blaze, creates a competitive game with continuous feedback on how well the developer adopts the improved practice.  The more they change their practices to adopt new techniques or tools, the higher they score in the game.

Would software developers writing serious software be motivated by this gamification concept? To answer this question, this work contributes an assessment of developers' response to game-like elements tied to their use of new tools and practices.  We assess 130 developers' response and relevant concerns about the idea of gamification through a web-based survey.  We piloted Blaze, our tool that introduces game elements focused on improving navigation practices into the developer's IDE.  With the pilot feedback we determine the acceptance, influence, and effect of deploying the Blaze tool in the pilot group.

The rest of this paper is organized as follows:  Section 2 reviews related work, Section 3 reviews results from the pre-study survey, Section 4 discusses the design considerations for developing the Blaze tool/game, Section 5 discusses the design of the Blaze pilot study,  Section 6 results from the pilot study and post-pilot survey, Section 7 covers threats to talidity, and Section 8 has our conclusions.
%This is precisely what we set out to study for this work. 
%
%Our first step was to challenge this question through a pre-study survey that queried both whether developers would be interested in gamifying their workspace and what aspects of that space would be most interesting to them.
%
%After confirming a good level of interest from the pre-study survey, we identified a set of practices and defined a game meeting expectations from the developers feedback.  The game, implemented in a Visual Studio add-in called Blaze, integrates into the Visual Studio workspace and establishes an anonymous competition between developers to improve their navigation practices. 
%
%During the pilot of Blaze we observed the variation in developer navigation practices and identified some potential reasons for this variation.  We conducted post-pilot survey and feedback sessions that evaluated developers' acceptance of game elements in the tool and their desires for additional functionality.

@@

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}. Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dbois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each others' rankings while the control group only sees their personal rankings.  Results though inconclusive, show competition increases the desired results.  In our study we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.

@@

Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?  We segmented results by country in Figure \ref{fig:gamification} and found gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicate developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.  Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.  Our results land on both sides of the 75\% mark resulting from the Saatchi and Saatchi study.

@@

\begin{figure}
	\includegraphics[width=3.4in]{ComfortWithSharing.pdf}
	\caption{Comfortable with sharing usage information}
	\label{fig:comfortwithsharing}
\end{figure}

We asked developers what would motivate them to use tools and practices and gave them 4 choices to rank.  The top motivator for using tools and practices is coworker recommendation with 75\% of respondents in Figure \ref{fig:toolandpracticemotivators}   ranking in 1st or 2nd place. The lowest was having badges posted on their social network profile where 33\% ranked this in 1st or 2nd place.   Badges also ranked low in motivations for taking more software engineering training.  We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.

@@

 To summarize, the high-level conclusions from each key question are:
\begin{itemize}[itemsep=0mm]
\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system 
\item Developers favor a team-oriented competition over individual awards
\item Awards for performing good quality development practices ranked highest
\item Over 90\% of respondents approved of sharing data with their team members
\end{itemize}
Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool focused on sharing developer data and rewarding developers for using good practices and tools.

@@

%\begin{enumerate}[itemsep=0mm]
%\item Define the practice we wish to measure compliance with
%\item Develop a mapping of the practice to actions that are captured by the blaze tool from the IDE
%\item Weight the actions according to their importance to the practice and assign points to the actions in the configuration file
%\item Develop the intervention method
%\item Determine the phases and length of the phase in the study according to baseline, intervention, and continuous feedback
%\item Select participants to deploy
%\item Define a pre-study questionnaire if one is necessary to collect associated information
%\item Define a post-study questionnaire if necessary to collect post study information
%\item Communicate with the participants about the study
%\item Execute the study and report results
%\end{enumerate}

%distinguish pilot from pre-study.  How does gamification apply?

@@

Another criteria is imposed by the instrument scope where we must limit ourselves to practices performed entirely in the IDE so they can be monitored using Blaze.  We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent CM submissions.

@@

Using these guidelines we assigned points to structured navigation commands which received 1 point each while unstructured navigation commands received 0 points.    We gave additional emphasis to using the Sando search tool by assigning it 10 points.  To create levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we applied an exponential curve based on points scored.  Users initially ``level up'' after  working a day or two; the next level may be achieved after a week or even a month following an exponential curve.  We designed the points and levels so above average developers can complete all levels during the study period.

\subsection{Competitive Elements}

By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition among teammates.    The leaderboard allows developers to see their own score in relation to the top 5 people on the team.  However, the identifier of the other participants was generated as 3 capital letters assigned based on their position on the leaderboard.  Not knowing who of your colleagues was ahead of you could be  reduce motivation, but this avoids one potential source of conflict in an intact team.

@@

Utilizing the continuous monitoring capabilities of Blaze, we designed the pilot to conduct a longitudinal study of the effect of Blaze on developers' navigation practices.  The Blaze tool revealed information about navigation in 3 stages of 1 week duration each.  A post-pilot survey gathered developers' opinions of Blaze and some confounding factors that could influence results.

@@

Part of the research objective was to determine whether we can achieve adoption of a practice with alternative interventions.  As previously discussed, we sought to provide Blaze to influence developers towards useing more structured navigation practices.  We constructed the interventions during the pilot to roll out in stages each week they use the tool.  This allowed us to establish a baseline and determine the changes as the pilot continued.

@@

\begin{figure}
	\centering
	\includegraphics[width=2.2in]{blazeWindow.png}
	\caption{Blaze Tool Window}
	\label{fig:blazeWindow}
\end{figure}

@@

%\subsubsection{Observations About Developers' Habits}
%
%We are interested in understanding whether developers are consistently using practices that make them inefficient.  One way to check that is to determine whether our measurements are associated with specific developers in the study, in other words, can we independently classify the edits performed by each developer.  If this is possible, then we can discuss whether developer have consistent ways of navigating and understanding for each edit.  Thus the hypothesis is developers have consistent ways of navigating and understanding code while they are making changes.
%
%To test this hypothesis we construct a classification tree using Weka \cite{Hall2009WEKA} with the J48 classifier.  Slicing the data and training the model using 10-fold cross-validation, the model identifies 53\% of the 647 edit sessions with the correct developer among 6 participants.  The tree uses nearly all available fields including time for all categories and time of day for edits in its attempt to associate sessions with their developer generating a Kappa statistic of 0.403  and square root of the relative squared error at 101\%.  
%
%While there is some consistency to how developers work, we cannot say that individual developers are always consistent in their work patterns.  Thus developers are somewhat flexible in the commands they use when navigating and making changes to code and may be more open to adopting new tools and techniques.\

@@

Using Weka\cite{Hall2009WEKA}, we performed an attribute selection process to identify the most significant attributes related to edit session duration.  The attribute search method we chose was Greedy Step-wise search selecting attributes based on their correlation with the target and stops adding attributes when the next attribute decreases the correlation.    The process identified the top 5 out of 15 possible categories as follows:
\begin{enumerate}[itemsep=0mm]
\item Unstructured Navigation
\item Debug - events from running the debugger
\item Edit - events that modify the code
\item Other Actions - other events in visual studio that we do not sub-classify
\item Build - events from running a build
\end{enumerate}

Using R\cite{Rcitation}, we constructed a linear regression model for Unstructured Navigation to determine the correlation with edit duration.  The linear model found a positive correlation with unstructured navigation with a p-value <0.001 on 1694 degrees of freedom.  The model had an Adjusted R-squared of 0.64 showing a good portion of the variation in edit duration is explained by the Unstructured Navigation category.

@@

Structured navigation events included: 
\begin{itemize}[itemsep=0mm]
\item Navigate To (Ctrl+,) is a fuzzy search interface that locates symbols matching your search terms.
\item Go To Definition (F12) brings up the code where the selected symbol is defined. 
\item View Call Hierarchy (Ctrl+K Ctrl+T) provides a two way analysis of a symbol's dependencies and uses. 
\item Class View (Ctrl+W, C) provides a browser and search function for classes and class hierarchy
\item Find All References provides a list of references for an object
\item Navigate to Event Handler brings up the event handler from the XAML editor
\item View Class Diagram generates a class diagram for the project
\item View Object Browser is a search tool for objects in the solution
\end{itemize}

@@

The developers using Blaze received feedback in the form of points given for using structured navigation commands and using the Sando search tool\cite{Shepherd2012Sando}.  Per the pilot design, the points display was disabled until the third week following installation to establish the efffect of instant feedback apart from other information.  The points display and the navigation arrow were enabled in the third week to test whether the instant feedback was successful in driving increased use of the targeted features. During the first two weeks shown in Figure \ref{fig:pointsbyweek}, developers were fairly consistent in their use patterns.  With the introduction of the points feedback, two developers' points accumulation spiked in week three due mostly to increased use of Sando.   We informed developers at the beginning of the pilot that Sando usage would receive 10 points and using built-in structured navigation commands would receive 1 point.

In these results we see an initial though not statistically significant effect from the points display in Blaze in Week three.  In week three, the developers could see their points feedback for the first time.  Two developers took more notice of this than others and used Sando more in the third week.  Their use of Sando continued in weeks 4 and 5 though with lower frequency.   The effect of points looks greater than navigation ratio during Week 3 because of the additional emphasis of the 10 point bonus given for Sando usage.  Developers in thir post-pilot feedback felt more guidance and more prompting included in the game would help guide them towards higher scores.

@@

\subsubsection{Direct Feedback Discussions}

@@

We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing. Thus categories like testing and debugging were affected by whether or not they used the virtual environment or their desktop for these activities.  Because those categories are not part of navigation, they are not discussed here. The presence of the virtual environment may explain how some developers have no activity recorded on some days.

@@

%Can we get a quote here?
One developer's build effort and the underlying build frequency was much higher than median as shown in Figure \ref{fig:developercomparison}.  We asked the developer to explain how building more frequently helps them. By doing more frequent build, (s)he is trying to ensure the correctness of the work completed so far. The developer feels like (s)he resolves the bugs faster with frequent builds because the code and logic is fresh in her memory. If they build less frequently, even though it reduces the build effort, there are more bugs each time and the developer often spends more time in identifying the problem area than fixing it. Hence the developer feels more frequent builds improves  productivity.

In another case, we observed a higher level of unstructured navigation than median and asked the developer to explain.  Even though they agree  structured navigation is better than unstructured navigation, (s)he prefers unstructured navigation in some specific instances. The developer is the author of the code and (s)he knows the code completely. ``I know that the line I need to fix is in XYZ method and it is 4 blocks down from where I am currently. It takes much less time for me to reach the fix location using unstructured navigation. If the code is not written by me, I will prefer structured navigation.'' This instance cautions us to gather information on aspects such as code familiarity and developer experience with the product when assessing the results of a navigation study.

%Though commonly held opinion of Microsoft's Clippy experience suggests this more invasive form of feedback is not always welcome, we shall list as future work providing more immediate suggestions on how developers can improve their navigation practices. 
Developers also reiterated how they wanted a different feedback experience from Blaze  such as immediate prompts to suggest navigation commands within the context of what they are working on.  Moving towards more immediate feedback, we shared a set of achievement badges tied to specific navigation commands  as we gave the more quantitative feedback to the developers.   Developers said badges would both inform and motivate them to use some specific tools or commands more than points alone. 
%  This contradicts a result from the pre-study survey where respondents rated badges as less important than team or individual competition in motivating them to improve.  The important difference in this case is badges serve as both a feedback mechanism and information showing which commands are important to use for improving structured navigation and their score.

@@

Measuring navigation ratio involved categorizing events into structured and unstructured navigation.  The categorization of events is specific to this study and documented herein, however, it may not be accurate to the intent of the operational definition of structure navigation depending on how the developer uses some commands like Find.  Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, we might consider something other than navigation (though it is classified as such for this paper).

@@

Threats to internal validity are those conditions that create alternative reasons for the conclusions, or confound the conclusion.  
In this study we may experience the Hawthorne effect where improvement occurs because we are seeking a change in a particular practice.  Specifically, it could affect the study because a presentation and demonstration to the pilot organization was required in advance of deployment.  By making the purpose of our measurement and contest known,  we may have triggered the participants to think about navigation commands they use before establishing a baseline.  Because we focus on the same participants with the same behavior, they may have learned how to demonstrate improvement at some time other than when presented with the information in Blaze.  This is mitigated by the data showing some participants did not improve over the course of the study while others showed they were attempting to improve.

>>>>

\title{Experiences Gamifying Developer Adoption\\ of Practices and Tools}
% 
% You need the command \numberofauthors to handle the 'placement
% and alignment' of the authors beneath the title.
%
% For aesthetic reasons, we recommend 'three authors at a time'
% i.e. three 'name/affiliation blocks' be placed beneath the title.
%
% NOTE: You are NOT restricted in how many 'rows' of
% "name/affiliations" may appear. We just ask that you restrict
% the number of 'columns' to three.
%
% Because of the available 'opening page real-estate'
% we ask you to refrain from putting more than six authors
% (two rows with three columns) beneath the article title.
% More than six makes the first-page appear very cluttered indeed.
%
% Use the \alignauthor commands to handle the names
% and affiliations for an 'aesthetic maximum' of six authors.
% Add names, affiliations, addresses for
% the seventh etc. author(s) as the argument for the
% \additionalauthors command.
% These 'additional authors' will be output/set for you
% without further effort on your part as the last section in
% the body of your article BEFORE References or any Appendices.

@@

As software development practices evolve, toolsmiths face the continuous challenge of communicating new practices and tools to the community.   We tested an idea with industrial software developers that adding game-like feedback to the development environment would improve adoption of tools and practices for code navigation.  We present results from a pre-study survey of 130 developers' opinions on gamification and motivation, data from a pilot by an intact team of six developers of a usage monitoring tool with game features tied to code navigation practices, and feedback collected in post-pilot interviews.  
Most developers expressed interest in gamification in our survey, though some have strong negative opinions.  Pilot results show that two of the six pilot developers adjusted their practices when presented with competitive game elements.
%We found that a software engineering game with competitive elements influences some developers towards changing their practices.
%We also learned that demonstrating significant improvement can be difficult and that developers want more proactive messages and recommendations on ways they can improve.
%
% In addition to training, discussing, and presenting software engineering practices and tools, we demonstrate a method to both motivate and monitor their adoption through a tool embedded in the Integrated Development Environment (IDE).  
%
%We determine through a survey of 130 developers the things we can do to motivate them to adopt 
%Blaze monitors developer practice and tool use over the long term and reports the lasting change a tool deployment or communication effort makes in everyday habits. 
%
% Embedding Blaze in the developer's work environment as a Visual Studio extension captures developer work patterns while giving feedback to users on their practices.   Blaze encourages everyone to use the new practices by assigning points to specific commands and tools.  
% With a rich data source, we evaluate developer practices, define metrics that give individual feedback, and create an environment for developers to see how they are doing compared with their teammates.  
%
%Data collected from the pilot group show a dichotomy of navigation and code understanding practices used by developers even the same developer in different situations.  We observed that the duration of an edit session is more correlated with categories of activity that benefit from improved practices and tools. The Blaze tool was well received by most participants, however, navigation practices were not significantly affected by the Blaze feedback mechanism.  We conclude that more proactive interventions are necessary to encourage developers to change their code navigation habits.
\end{abstract}

@@

Sascha is a developer with years of experience maintaining code, who knows how to navigate through code fairly well.  Sascha heard about some other navigation tools, but kept using the built-in features of the Visual Studio IDE mainly out of habit.  When Sascha's company started a game around code navigation in Visual Studio, he became interested in being a part of the fun.  Sascha tried the code navigation tools and practices that were part of the game and soon found himself at the top of the game's leaderboard.  After learning the new practices Sascha spends less time navigating code to accomplish his maintenance tasks.

Like Sascha, software developers who maintain code face challenges of navigating an existing code base to incorporate changes as they build new features or fix bugs.  Toolsmiths respond to these challenges by creating new tools for assisting developers through difficult activities like code navigation.  However, as new practices and tools become available, developers may be motivated to start using them unless they interact with a peer developer who promotes them\cite{wbsnipes:Hill2011Peer}.    Of course they are intrinsically motivated to improve, but urgent demands interfere with experimenting using new tools and they become less efficient than otherwise possible.  Thus we need to both communicate with and motivate developers like Sascha to start using new tools by showing them how they benefit from using the new tools.

One option for motivating people to adopt new practices is to apply elements of game mechanics or ``gamify'' the practice. Gamification is the incorporation of game elements into an activity that people do not typically consider a game\cite{2013Oxford}.  We envisioned a tool in our prior work that would motivate developers by providing constant feedback and allowing them to compare their improvements to how their peers are doing  \cite{Snipes2013Towards}.  This tool, called Blaze, creates a competitive game with continuous feedback on how well the developer adopts the improved practice.  The more they change their practices to adopt new techniques or tools, the higher they score in the game.

Would software developers writing serious software be motivated by this gamification concept? To answer this question, this work contributes an assessment of developers' response to game-like elements tied to their use of new tools and practices.  We assess 130 developers' response and relevant concerns about the idea of gamification through a web-based survey.  We piloted Blaze, our tool that introduces game elements focused on improving navigation practices into the developer's IDE.  Developers' reactions to gamification were mostly positive with a few detractors, and results of piloting show that game elements influence some developers but have no effect on others' use of tools and practices.

The rest of this paper is organized as follows:  Section 2 reviews related work, Section 3 reviews results from the pre-study survey, Section 4 discusses the design considerations for developing the Blaze tool/game, Section 5 discusses the design of the Blaze pilot study,  Section 6 results from the pilot study and post-pilot survey, Section 7 covers threats to validity, and Section 8 has our conclusions.

@@

Dubois and Tamburrelli discuss theoretical concepts behind applying gamification to software development \cite{Dubois2013Understanding}. Their approach is to consider how gamification could be applied throughout the software development life-cycle to motivate many practices in different phases.  They provide guidance on assessing gamification in the software development domain including analyzing the actors and game mechanics, defining integration with existing tools and processes, and designing the evaluation of results.  Dubois and Tamburrelli demonstrate this guidance using a study of students working on a class project.  Defining a control group and subject group of students, both groups receive the same feedback about their code comments, test coverage, code size, and static analysis rule violations.  The subject group can see each others' rankings while the control group only sees their personal rankings.  Results though inconclusive, show competition increases the desired results.  In our study we provide participants with both individual feedback and anonymized competitive feedback and seek to observe a change in their behavior.

@@

Using a question from the Saatchi and Saatchi survey \cite{wbsnipes:SaatchiGameification} as directly as possible, we asked developers at ABB ``How interested they would be in working for a company that incorporated some aspects of games into software engineering tools as a way to increase productivity in the workplace''?  We segmented results by country in Figure \ref{fig:gamification} and found gamification of software engineering is more interesting to developers in India, Poland, and Finland than in other major countries.  The responses indicate developers were at least somewhat interested is 75-88\% In India, Poland, and Finland.  For the US, Sweden, and Switzerland the responses indicate 60-65\% are at least somewhat interested.

Comments from participants reflected a more interesting controversial reaction than these measurements might indicate.  A developer from India said ``Gamification is a juvenile way of getting programmers to do something that you want.''  A different respondent expressed a concern that ``Gamification of best practices can have nasty side effects.''  Another participant said ``there is already high a level of competition between software development teams..'' indicating they are under considerable pressured to get the work done.  Another developer reacted with a statement that identifies the kind of intrinsic motivation we hope for ``I am using practices and tools because I want to do my job well',' indicating that gamification was unnecessary to motivate people to improve.

On the positive side, a developer from Poland said ``In a previous work place we had a 'game' between developers - scoring broken/unbroken builds and new/passed/failed tests. It was helping to improve quality and It worked well. Recommended especially for junior developers. ''  Other comments indicated competition or points would be acceptable ``as long as it doesn't interrupt or slow down my work. ''

Because we nearly identically reused their original question, we compare the overall response from our survey to the Saatchi and Saatchi survey\cite{wbsnipes:SaatchiGameification} upon which this question was based.  Our results land on both sides of the 75\% mark resulting from the Saatchi and Saatchi study.

@@

\begin{figure}
	\includegraphics[width=3.4in]{ComfortWithSharing.pdf}
	\caption{Comfort with sharing usage information}
	\label{fig:comfortwithsharing}
\end{figure}

We asked developers what would motivate them to use tools and practices and gave them four choices to rank.  The top motivator for using tools and practices is coworker recommendation with 75\% of respondents in Figure \ref{fig:toolandpracticemotivators}   ranking in 1st or 2nd place.  Of several comments, a developer in Sweden captured the following scene, ``[The] most motivating [thing] is when a new practice / tool is discussed during a continuous improvement meeting or even coffee break, where one co-developer tells about the benefits and there is a joint discussion on where it (tool or practice) could be best applied.''

The team goal ranked second behind coworker recommendation.  This response indicated that developers would be motivated to use new tools and practices if that helped their team in a competion between development teams at ABB.  The result shows that team competition would be first or second on the list for  47\% of developers to motivate their use of practices or tools.

Just behind team competition is management mandate where developers would be asked to follow specific practices by the organization.  The lowest was having badges posted on their social network profile where 32\% ranked this in 1st or 2nd place.   Badges and management mandate received the most negative written comments in the survey.  A developer from the United States captured the lower opinion of badges that several participants expressed ``I am motivated because ABB allows me to choose the tools that best work for me. A badge will not really do anything for me.''  Badges also ranked low in motivations for taking more software engineering training.

We conclude that a facility to share coworker recommendations on tools and practices has the best opportunity to increase their adoption.

@@

 To summarize, the high-level conclusions from each key question are:
\begin{itemize}[itemsep=0mm]
\item 95\% of developers responding would try tools and practices suggested by an automated recommendation system 
\item Developers are motivated by collaboration and team goals more than mandates and individual awards
\item Awards for performing good quality development practices ranked highest
\item Over 90\% of respondents approved of sharing data with their team members
\end{itemize}
Thus, this pre-study survey provided a strong positive indicator for going forward with a software engineering tool focused on sharing developer data and rewarding developers for using good practices and tools.

@@

@@

Another criteria imposed by the instrument limits us to practices performed entirely in the IDE so they can be monitored using Blaze.  We identified potential practices of test-driven development, debugging practices, navigation practices,  eliminating static analysis bugs, and frequent CM submissions.

@@

Using these guidelines we assigned points to structured navigation commands which received one point each while unstructured navigation commands received zero points.    We gave additional emphasis to using the Sando search tool by assigning it ten points.  To create levels, following the guidance from Hamari \cite{wbsnipes:Hamari2011Framework}, we applied an exponential curve based on points scored.  Users initially ``level up'' after  working a day or two; the next level may be achieved after a week or even a month following an exponential curve.  We designed the points and levels so above average developers can complete all levels during the study period.

By selecting an intact team, we hope to leverage aspects of motivation such as the Leaderboard to spur feelings of competition between teammates.    The leaderboard allows developers to see their own score in relation to the top five people on the team.  However, the identifier of the other participants was generated as three capital letters assigned based on their position on the leaderboard.  Not knowing who of your colleagues was ahead of you could be  reduce motivation, but this avoids one potential source of conflict in an intact team.

@@

Utilizing the continuous monitoring capabilities of Blaze, we designed the pilot to conduct a longitudinal study of the effect of Blaze on developers' navigation practices.  The Blaze tool revealed information about navigation in three stages of one week duration each.  A post-pilot survey gathered developers' opinions of Blaze and some confounding factors that could influence results.

@@

Part of the research objective was to determine whether we can achieve adoption of a practice with alternative interventions.  As previously discussed, we sought to provide Blaze to influence developers towards using more structured navigation practices.  We constructed the interventions during the pilot to roll out in stages each week they use the tool.  This allowed us to establish a baseline and determine the changes as the pilot continued.

@@

\begin{figure}
	\centering
	\includegraphics[width=1.6in]{blazeWindow.png}
\includegraphics[width=1.6in]{blazeLeaderboard.png}
	\caption{Blaze Tool Window}
	\label{fig:blazeWindow}
\end{figure}

@@

@@

Using Weka\cite{Hall2009WEKA}, we performed an attribute selection process to identify the most significant attributes related to edit session duration.  The attribute search method we chose was Greedy Step-wise search selecting attributes based on their correlation with the target and stops adding attributes when the next attribute decreases the correlation.    The top 5 out of the 15 possible categories ranked by this process are as follows:
\begin{enumerate}[itemsep=0mm]
\item Unstructured Navigation
\item Debug - events from running the debugger
\item Edit - events that modify the code
\item Other Actions - other events in visual studio that we do not sub-classify
\item Build - events from running a build
\end{enumerate}

Using R\cite{Rcitation}, we constructed a linear regression model for Unstructured Navigation to determine the correlation with edit duration.  The linear model found a positive correlation with unstructured navigation with a p-value <.001 on 1694 degrees of freedom.  The model had an Adjusted R-squared of 0.64 showing a good portion of the variation in edit duration is explained by the Unstructured Navigation category.

@@

Structured navigation events included: 
\begin{itemize}[itemsep=0mm]
\item Navigate To (Ctrl+,) is a fuzzy search interface that lists identifiers matching the selected string 
\item Go To Definition (F12) brings up the code that defines the selected identifier 
\item View Call Hierarchy (Ctrl+K Ctrl+T) provides a two way analysis of an identifier's dependencies and uses
\item Class View (Ctrl+W, C) provides a browser and search function for classes and class hierarchy
\item Find All References provides a list of lines that reference an identifier
\item Navigate to Event Handler in the XAML editor shows the event handler for an object
\item View Class Diagram generates a class diagram 
\item View Object Browser is a search tool and browser
\end{itemize}

@@

The developers using Blaze received feedback in the form of points given for using structured navigation commands and using the Sando search tool\cite{Shepherd2012Sando}.  Per the pilot design, the points display was disabled until the third week following installation to establish the effect of instant feedback apart from other information.  The points display and the navigation arrow were enabled in the third week to test whether the instant feedback was successful in driving increased use of the targeted features. During the first two weeks shown in Figure \ref{fig:pointsbyweek}, developers were fairly consistent in their use patterns.  With the introduction of the points feedback, two developers' points accumulation spiked in week three due mostly to increased use of Sando.   We informed developers at the beginning of the pilot that Sando usage would receive ten points and using built-in structured navigation commands would receive one point.

In these results we see an initial though not statistically significant effect from the points display in Blaze in Week three.  In week three, the developers could see their points feedback for the first time.  Two developers took notice of the points and used Sando more in the third week.  Their use of Sando continued at a lower level in weeks four and five.   The effect of points feedback is greater than navigation ratio during week three because of the additional emphasis of the ten point bonus given for Sando usage.  Results from our post-pilot survey where at least half the developers said the points and leaderboard influenced them "a lot" also support our conclusion that some though not all developers were influenced by the game elements in Blaze.

@@

\subsubsection{Post-Pilot Feedback Discussions}

@@

We discovered during these feedback sessions that developers did not spend the entire time in their desktop environment because they had a virtual machine environment available for testing. Thus categories like testing and debugging were affected by whether they used the virtual environment or their desktop for these activities.  Because those categories are not part of navigation, they are not discussed here. The presence of the virtual environment may explain how some developers have no activity recorded on some days.

@@

One developer's build effort and the underlying build frequency was much higher than median as shown in Figure \ref{fig:developercomparison}.  We asked the developer to explain how building more frequently helps them. By doing more frequent build, (s)he is trying to ensure the correctness of the work completed so far. The developer feels like (s)he resolves the bugs faster with frequent builds because the code and logic is fresh in her memory. If they build less frequently, even though it reduces the build effort, there are more bugs each time and the developer often spends more time in identifying the problem area than fixing it. Hence the developer feels more frequent builds improves her productivity.

In another case, we observed a higher level of unstructured navigation than median and asked the developer to explain.  Even though they agree  structured navigation is better than unstructured navigation, (s)he prefers unstructured navigation in some specific instances. The developer is the author of the code and (s)he knows the code completely. ``I know that the line I need to fix is in XYZ method and it is four blocks down from where I am currently. It takes much less time for me to reach the fix location using unstructured navigation. If the code is not written by me, I will prefer structured navigation.'' This instance cautions us to gather information on aspects such as code familiarity and developer experience with the product when assessing the results of a navigation study.

Developers also reiterated how they expected a different experience from Blaze  such as immediate prompts to suggest navigation commands within the context of what they are working on.  This contradicts some developer comments from the pre-study survey that indicated interruptions from a game tool would be distracting and perhaps unwelcome.

In the feedback, we shared a set of achievement badges tied to specific navigation commands with developers.   Developers said badges would both inform and motivate them to use some specific tools or commands.   This contradicts a result from the pre-study survey where respondents rated badges last in their influence motivating them to improve.  The difference in this opinion may result from the context where developers, when presented with badges, see how badges serve as both a feedback mechanism and information on the commands for improving structured navigation and their score.

@@

Measuring navigation ratio involved categorizing events into structured and unstructured navigation.  The categorization of events is specific to this study and documented herein, however, it may not be accurate to the intent of the operational definition of structured navigation depending on how the developer uses some commands like Find.  Another factor affecting this measurement may occur when developers repeat a command such as hitting the down arrow key many times, that may be something other than navigation (though it is classified as such for this paper).

@@

Threats to internal validity are those conditions that create alternative reasons for the conclusions, or confound the conclusion.  
In this study we may experience the Hawthorne effect where improvement occurs because we are seeking a change in a particular practice.  Specifically, it could affect the study because a presentation and demonstration to the pilot organization was required in advance of deployment.  By making the purpose of our measurement and contest known,  we may have triggered the participants to think about navigation commands they use before establishing a baseline.   This threat is mitigated by the data showing four participants did not improve over the course of the study while two demonstrated some improvement.

Participants in the pilot were not randomly selected, they are all volunteers from one intact team chosen due to their advanced practices and understanding of metrics.  Therefore the participants may be predisposed to have an interest in the kind of tool and methodology we are piloting and have bias towards a favorable opinion.

The intact team may be more familiar with the code they maintain thus be less likely to change their navigation practices.  The post-pilot survey responses confirmed developers are less inclined to use structured navigation when they work in code they are very familiar with.

Threats to external validity include the study was conducted in one development location of one company, other companies and locations may have different cultures incompatible with this type of monitoring and gamification activity.

 Threats to generalizable results exist because we conducted the pilot of the tool with six developers thus the population size is too small to draw statistically significant inferences from.
